---
title: "Data Exploration on 2021 England Census Data"
author: "Saola Gbenga Ifeoluwa"
format:
  html:
    code-fold: true
    code-link: true
    toc: true
    toc-location: left
    toc-expand: true
    number-sections: true
    
bibliography: references.bib
editor: visual
---

```{=html}
<style>
h1 {
  font-size: 20pt;
}
</style>
```
```{=html}
<style>
h2 {
  color: #424858;
  font-size: 16pt;
}
</style>
```
# Loading R packages

I used tidyverse documentation [@tidyverse2019] for data cleaning and R Programming code for exploration and visualization [@rsoftware] and Quarto [@quarto2022] documentations for markdown.

```{r}
#| label: load-packages
#| include: true
#| eval: true
#| results: hide
library(ggplot2)
library(patchwork)
library(tidyverse)
```

# Reading dataset into RStudio

```{r}
data <- read.csv("C:/Users/gsaola/Downloads/data-1.csv")
head(data)
```

# Data preprocessing & feature engineering

## Handling mising values

Missing values spotted in the data are below: Mar_Sat=6,144 \| INC=6,173 \| Highest.Ed=1,123

```{r}
clean_data <- data[complete.cases(data), ]
```

## Categorize Age data into Age_Group

```{r}

age_breaks <- c(-Inf, 10, 20, 30, 40, 50, 60, Inf)
age_labels <- c("<10", "10-19", "20-29", "30-39", "40-49", "50-59", "60+")
clean_data$Age_Group <- cut(clean_data$Age, breaks = age_breaks, labels = age_labels, right = FALSE)
```

## Renaming columns

```{r}
new_column_names <- c("Marital_Status", "Income", "Gender", "all_rooms_behind_door", "Ethnicity", "Highest_education")
colnames(clean_data)[c(4, 5, 6, 7, 8,9)] <- new_column_names
```

## Renaming column values of gender

0:Female \| 1:Male

```{r}
clean_data$Gender <- ifelse(clean_data$Gender == 0, "Female", "Male")
clean_data$all_rooms_behind_door <- ifelse(clean_data$all_rooms_behind_door == 0, "Yes", "No")
```

## Taking a peek at the Cleaned data

```{r}
head(clean_data)
```

# Data Visualization

## Demographic Overview - Population Pyramid

The stacked bar below shows the distribution of age by gender for guidance on product curation and marketing strategies tailored to specific age groups.

```{r}
ggplot(clean_data, aes(x = Age_Group, fill = Gender)) +
  geom_bar(stat = "count", position = "dodge") +
  coord_flip() +
  labs(title = "Demographic Overview - Age Distribution",
       x = "Count",
       y = "Age Group")
```

## Marital Status Composition - Pie Chart

The pie chart below displays the distribution of individuals across different marital statuses, offering a quick overview of family structures.

```{r}
ggplot(clean_data, aes(x = "", fill = Marital_Status)) +
  geom_bar(width = 1, stat = "count") +
  coord_polar(theta = "y") +
  labs(title = "Marital Status Composition",
       fill = "Marital Status") +
  theme_void()
```

## Income Distribution - Histogram

The graph below shows income distribution, explaining how pricing strategies and products can be affected by customer income.

```{r}
ggplot(clean_data, aes(x = Income)) +
  geom_histogram(binwidth = 5000, fill = "skyblue", color = "black") +
  labs(title = "Income Distribution",
       x = "Annual Income (in pounds)",
       y = "Count")
```

## Privacy Preferences - Stacked Bar Chart

The chart below shows the privacy preference of individuals in the data which can be factored into product design.

```{r}
ggplot(clean_data, aes(x = factor(all_rooms_behind_door), fill = factor(all_rooms_behind_door))) +
  geom_bar() +
  labs(title = "Privacy Preferences",
       x = "Response to Question H8",
       y = "Count")
```

## Education distribution

```{r}
plot5 <- clean_data %>%
  count(Highest_education) %>%
  ggplot(aes(x = Highest_education, y = n)) +
  geom_bar(stat = "identity") +
  ggtitle("Education Distribution") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot5
```

## Age Distribution

```{r}
plot6 <- clean_data %>%
  count(Age_Group) %>%
  ggplot(aes(x = Age_Group, y = n)) +
  geom_bar(stat = "identity") +
  ggtitle("Age Group Distribution") +
  theme(axis.text.x = element_text(size = 15))

plot6
```

## Education levels by gender

```{r}
plot7 <- ggplot(clean_data, aes(x = Highest_education, fill = Gender)) +
  geom_bar(position = "dodge", show.legend = TRUE) +
  labs(title = "Distribution of Education Levels by Gender")

plot7  + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Income distribution by family structure and age

Below is the income distribution by family structure, providing further insight to which to target for family products and premium services.

```{r}
ggplot(clean_data, aes(x = Age, y = Income, color = Marital_Status)) +
  geom_point(size = 3) +  # Adjust the size as needed
  theme(
    plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")  # Adjust the margins as needed
  ) +
  labs(title = "Scatter Plot of Age vs Income by Marital Status")
```

# Conclusion

Older people are more than the young ones and therefore, it does not fully represent the population of customers that can be targeted for this business.

Secondly, majority of the people in the dataset are low income earners, if further details on location could be provided, further analysis could be carried out to know if a new area should be considered for the premium products

# Recommendation

Market research should be carried out to know the preferences of customers when it comes to home products and services that best suit their age group, income and privacy.

A well distributed sample of data should be provided, such that would represent the entire population.

```{references}
```
